---
title: "Class 11 Introduction to Genome Informatics Lab"
author: 'Justin Robinson (PID: A16307501)'
date: "2024-05-07"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Population Scale Analysis Homework

> Q13: Read this file into R and determine the sample size for each genotype and their corresponding median expression levels for each of these genotypes.

```{r}
geno_table <- read.table(url("https://bioboot.github.io/bggn213_W19/class-material/rs8067378_ENSG00000172057.6.txt"))
```

```{r}
sum(geno_table$geno == "A/G")
sum(geno_table$geno == "A/A")
sum(geno_table$geno == "G/G")
```

Stats for the A/A group:
```{r}
library(dplyr)

AA_geno <- geno_table %>%
  filter(geno == "A/A")

summary(AA_geno)
```
The median expression of ORMDL3 for the A/A genotype is 31.25


Stats for the A/G group:
```{r}
AG_geno <- geno_table %>%
  filter(geno == "A/G")

summary(AG_geno)
```
For the A/G genotype it is 25.065


Stats for the G/G group:
```{r}
GG_geno <- geno_table %>%
  filter(geno == "G/G")

summary(GG_geno)
```
For the G/G genotype it is 20.074


> Q14: Generate a boxplot with a box per genotype, what could you infer from the relative expression value between A/A and G/G displayed in this plot? Does the SNP effect the expression of ORMDL3?

```{r}
boxplot(exp ~ geno, data = geno_table)
```
Based on the above boxplot, it would seem that the A/A genotype has a higher expression of the ORMDL3 gene. A/G is second and G/G is last in terms of ORMDL3 expression.
