---
title: "Class 8: Breast Cancer Mini Project"
format: html
---

## About

In today's lab we will work with fine needle aspiration (FNA) of a breast mass data from the University of Wisconsin.

## Data Import

```{r}
wisc.df <- read.csv("WisconsinCancer.csv", row.names = 1)
head(wisc.df)
```


> Q1. How many patients/individuals/samples are in this dataset?

```{r}
nrow(wisc.df)
```

> Q2. How many of the observations have a malignant diagnosis?

```{r}
sum(wisc.df$diagnosis == "M")
```
or do it this way
```{r}
table(wisc.df$diagnosis)
```

> Q3. How many variables/features in the data are suffixed with _mean?

```{r}
ncol(wisc.df)
```

```{r}
inds <- grep("_mean", colnames(wisc.df))
length(inds)
```

## Initial Analysis

Before analysis I want to take out the expert diagnoses column (aka the answer) from the dataset

```{r}
diagnosis <- as.factor(wisc.df$diagnosis)
diagnosis
```

```{r}
wisc.data <- wisc.df[,-1]
```


## Clustering

We can try kmeans() clustering first

```{r}
km <- kmeans(wisc.data, centers=2)
```

```{r}
table(km$cluster)
```

Corss-table

```{r}
table(km$cluster, diagnosis)
```

lets try `hclust()` the key input required for `hclust()` is a distance matrix as produced by the `dist()` function

```{r}
hc <- hclust(dist(wisc.data))
```

```{r}
plot(hc)
```

## PCA

We can look at the sd of each column

```{r}
round( apply(wisc.data, 2, sd) )
```

yes, we need to scale. We will run `prcomp()` with `scale=TRUE` 

```{r}
wisc.pr <- prcomp(wisc.data, scale=TRUE)
summary(wisc.pr)
```

Generate our main PCA plot (score plot, PC1 vs PC2 plot)...

```{r}
library(ggplot2)

res <- as.data.frame(wisc.pr$x)

```

```{r}
ggplot(res) + 
  aes(PC1, PC2, col=diagnosis) +
  geom_point()
```


## Combining methods

Clustering on PCA results

Using the minimum number of principal components required to describe at least 90% of the variability in the data, create a hierarchical clustering model with the linkage method="ward.D2". We use Ward’s criterion here because it is based on multidimensional variance like principal components analysis. Assign the results to wisc.pr.hclust.

```{r}
d <- dist( wisc.pr$x[,1:3] )
hc <- hclust(d, method="ward.D2")
plot(hc)


```

To get my clustering result/membership vector I need to "cut" the tree with the `cutree()` function

```{r}
grps <- cutree(hc, k=2)
```

. Q. How many patients are in each cluster group?

```{r}
table(grps)
```

##Prediction


We can use our PCA result (model) to do prediction, that is take new unseen data and project it onto our new PC variables.

```{r}
url <- "https://tinyurl.com/new-samples-CSV"
new <- read.csv(url)
npc <- predict(wisc.pr, newdata=new)
npc
```
> Q17. Which of your analysis procedures resulted in a clustering model with the best specificity? How about sensitivity?

The first scoreplot of PC1 vs PC2 had higher sensitivity because there were many dots that overlapped and could have shown either benign or malignant. The hierarchical clustering graph showed a clear divide so had higher specificity

```{r}
plot(res$PC1, res$PC2, col=grps)
points(npc[,1], npc[,2], col="blue", pch=16, cex=3)
text(npc[,1], npc[,2], labels=c(1,2), col="white")
```

> Q18. Which of these new patients should we prioritize for follow up based on your results?

Definitely 2, as they are within the area of the graph associated with malignancy

# Summary

Principal Component Analysis (PCA) is a super useful method for analyzing large datasets. It works bu finding new variables (PCs) that capture the most variance from the original variable in our dataset.